Discrete Quantum Theories and Computing

### 2018-02-24 dissertation.tex
dissertation.lyx "3.2 Modal Quantum Theory" Updating...

\documentclass[twoside]{iuphd}
\usepackage[T1,T2A]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[ backend=biber]{biblatex}
\addbibresource{prop.bib}
\ExecuteBibliographyOptions{sorting=none,maxbibnames=10,doi=false,isbn=false,url=false}
\DeclareUnicodeCharacter{0229}{\c{e}}
\DeclareUnicodeCharacter{00A0}{~}
\usepackage{mmap}
\usepackage{comment}
%\usepackage[basic]{complexity}
\newcommand{\lang}[1]{{\ensuremath{\mathsf{#1}}}}
\newcommand{\newlang}[2]{\newcommand{#1}{\lang{#2}}}
\newlang{\usat}{UNIQUE-SAT}
\usepackage[all]{xy}
\usepackage{framed}
\usepackage[braket,qm]{qcircuit}
\usepackage{mathrsfs}
\usepackage{amssymb}
\usepackage{pdfpages}
\usepackage{setspace}

\newcommand{\says}[3]{\begin{framed}\begin{minipage}{0.9\linewidth}\color{#1}{#2 says: #3}\end{minipage}\end{framed}}
\newcommand{\muB}{\ensuremath{\mu^{B}}}
\newcommand{\events}{\ensuremath{\mathcal{E}}}
\newcommand{\set}[2]{\ensuremath{\left\{ {#1}\mathrel{}\middle|\mathrel{}{#2}\right\} }}
\newcommand{\gmult}{*}
\newcommand{\Fpx}[1]{\mathbb{F}_{#1}}
\newcommand{\Fq}{\mathbb{F}_{q}}
\newcommand{\ff}[1]{\mathbb{F}_{#1}}

\begin{document}
%data for Title Page
\title{Discrete Quantum Theories and Computing}
\author{\large{Yu-Tsung Tai}}
\date{\today}
\department{Department of Mathematics}
\department{Department of Computer Science}
\maketitle

%data for Acceptance Page
\committeeMember{Amr A. Sabry, PhD}
\committeeMember{Dylan Paul Thurston, PhD}
\committeeMember{Gerardo Ortiz, PhD}
\committeeMember{Andrew J. Hanson, PhD}
\committeeMember{Shouhong Wang, PhD}
\defensedate{Defense Date}
\acceptancepage

@@

\section{Geometrical Structure of States}

@@

@@

The characteristic of a field is the least positive integer~$m$
such that $m=1+1+1+\cdots+1=0$, and if no such $m$ exists we say
that the field has characteristic zero (which is the case for $\mathbb{R}$
for example). It turns out that if the characteristic is non-zero
it must be a prime~$p$. For every prime~$p$ and positive integer~$r$
there is a finite field~$\Fpx{p^{r}}$ of size $q=p^{r}$ and characteristic~$p$,
which is unique up to field isomorphism \cite{Artin1991,DummitFoote2004}.
The exponent $r$ is known as the \emph{degree} of the field over
its prime subfield\footnote{ Fields $\ff{q}$ where $q$ is a power of a prime $p$, i.e., $q=p^{r}$,
are known as Galois fields.}~\cite{GT.ref}. If the characteristic $p$ is an arbitrary prime
number, we call the field \emph{unrestricted}.

@@

>>>>

%% LyX 2.2.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[twoside]{iuphd}
\usepackage[T1,T2A]{fontenc}
\usepackage[utf8]{inputenc}
\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}
\usepackage{verbatim}
\usepackage{mathrsfs}
\usepackage{pdfpages}
\usepackage{amsmath}
\usepackage{amssymb}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage[ backend=biber]{biblatex}
\addbibresource{prop.bib}
\ExecuteBibliographyOptions{sorting=none,maxbibnames=10,doi=false,isbn=false,url=false}
\DeclareUnicodeCharacter{0229}{\c{e}}
\DeclareUnicodeCharacter{00A0}{~}
\usepackage{mmap}
\usepackage{comment}

%\usepackage[basic]{complexity}
\newcommand{\lang}[1]{{\ensuremath{\mathsf{#1}}}}
\newcommand{\newlang}[2]{\newcommand{#1}{\lang{#2}}}
\newlang{\usat}{UNIQUE-SAT}

\usepackage[all]{xy}
\usepackage{framed}
\usepackage[braket,qm]{qcircuit}
\usepackage{mathrsfs}
\usepackage{setspace}

\newcommand{\says}[3]{\begin{framed}\begin{minipage}{0.9\linewidth}\color{#1}{#2 says: #3}\end{minipage}\end{framed}}
\newcommand{\muB}{\ensuremath{\mu^{B}}}
\newcommand{\events}{\ensuremath{\mathcal{E}}}
\newcommand{\set}[2]{\ensuremath{\left\{ {#1}\mathrel{}\middle|\mathrel{}{#2}\right\} }}
\newcommand{\gmult}{*}
\newcommand{\Fpx}[1]{\mathbb{F}_{{#1}}}
\newcommand{\Fq}{\Fpx{q}}
\newcommand{\ff}[1]{\Fpx{#1}}
\newcommand{\ffzx}[2]{\Fpx{#1}^{{#2}\;*}}
\newcommand{\ffx}[2]{\Fpx{#1}^{{#2}}}
\newcommand{\ffzd}[1]{\ffzx{#1}{d}}
\newcommand{\ffd}[1]{\ffx{#1}{d}}
\newcommand{\VVec}[1]{\vec{\bf #1}}
\newcommand{\mathComplex}{\mathbb{C}}
\newcommand{\braket}[2]{\ip{#1}{#2}}

\department{Department of Mathematics}
\department{Department of Computer Science}

\makeatother

\begin{document}
%data for Title Page

\title{Discrete Quantum Theories and Computing}

\author{{Yu-Tsung Tai}}

\date{\today}

\maketitle
%data for Acceptance Page
\committeeMember{Amr A. Sabry, PhD} \committeeMember{Dylan Paul
Thurston, PhD} \committeeMember{Gerardo Ortiz, PhD} \committeeMember{Andrew
J. Hanson, PhD} \committeeMember{Shouhong Wang, PhD} \defensedate{Defense
Date} \acceptancepage

@@

\section{Geometrical Structure of States\label{CQC1qubitBloch.sec}}

Conventional quantum theory (CQT) is described by the following: 
\begin{enumerate}
\item[(i)] $D=2^{n}$ orthonormal basis vectors of an $n$-qubit state, 
\item[(ii)] the normalized $D$ complex probability amplitude coefficients describing
the contribution of each basis vector, 
\item[(iii)] a set of probability-conserving unitary matrix operators that suffice
to describe all required state transformations of a quantum circuit, 
\item[(iv)] and a measurement framework. 
\end{enumerate}
We remark that there are many things that are assumed in CQT, such
as the absence of zero norm states for non-zero vectors, and the decomposition
of complex amplitudes into a pair of ordinary real numbers. One also
typically assumes the existence of a Hilbert space with an orthonormal
basis, allowing us to write $n$-qubit \textit{pure} states in general
as Hilbert space vectors with an Hermitian inner product: 
\begin{eqnarray}
\ket{\Psi} & = & \sum_{i=0}^{D-1}\alpha_{i}\ket{i}\,.
\end{eqnarray}
Here $\alpha_{i}\in\mathComplex$ are complex probability amplitudes,
$\VVec{\alpha}\in\mathComplex^{D}$, and the $\{\ket{i}\}$ is an
orthonormal basis of states obeying 
\begin{equation}
\braket{i}{k}=\delta_{ik}\,,
\end{equation}
where $\bra{i}$ is called the dual vector of $\ket{i}$.

The meaning of this is that any state $\ket{\Phi}=\sum_{i=0}^{D-1}\beta_{i}\ket{i}$
can be projected onto another state $\ket{\Psi}$ by writing 
\begin{eqnarray}
\braket{\Phi}{\Psi} & = & \sum_{i=0}^{D-1}\beta_{i}^{*}\alpha_{i}\,,
\end{eqnarray}
thus quantifying the proximity of the two states. (Here $^{*}$ denotes
complex conjugation.) This is one of many properties we take for granted
in continuum quantum mechanics that challenge us in defining a discrete
quantum geometry.

In this paper, we focus on the discrete geometric issues raised by
the properties (i) and (ii) given above for CQT, and leave for another
time the important issues of (iii), (iv), and such conundrums as probabilities,
zero norms, and dynamics in the theory of DQT. To facilitate the transition
to DQT carried out in later sections, we concern ourselves first with
the properties of the simplest possible abstract state object in CQT,
the single qubit state.

\begin{comment}

\subsection{The single qubit problem}

A single qubit already provides access to a wealth of geometric information
and context. We write the single qubit state as 
\begin{equation}
\begin{array}{lr}
\ket{\psi_{1}}=\alpha_{0}\ket{0}+\alpha_{1}\ket{1}\  & \ \alpha_{0},\alpha_{1}\in\C\end{array}\ .%%
%%\alpha_{1}{array}
\label{A1.eq}
\end{equation}

A convenience for probability calculations and a necessity for computing
relative state properties is the normalization condition 
\begin{equation}
\|\psi_{1}\|^{2}=|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=\alpha_{0}^{*}\alpha_{0}+\alpha_{1}^{*}\alpha_{1}=1\ ,\label{A2.eq}
\end{equation}
which identifies $\alpha_{0}$ and $\alpha_{1}$ as (complex) probability
amplitudes and implies the conservation of probability in the closed
world spanned by $\{\ket{0},\ket{1}\}$. Note that we distinguish
for future use the \textit{norm\/} $\|\cdot\|$ of a vector from
the \textit{modulus\/} $|\cdot|$ of a complex number. Continuing,
we see that if we want only the irreducible state descriptions, we
must supplement the process of computing \Eq{A2.eq} by finding a
way to remove the distinction between states that differ only by an
overall phase transformation $e^{i\phi}$, that is, 
\begin{equation}
(\alpha_{0},\alpha_{1})\sim(e^{i\phi}\alpha_{0},e^{i\phi}\alpha_{1})\ .\label{A3.eq}
\end{equation}
This can be accomplished by the Hopf fibration, which we can write
down as follows: let 
\begin{equation}
\begin{array}{cc}
\alpha_{0}=x_{0}+iy_{0}, & \alpha_{1}=x_{1}+iy_{1}\end{array}\ .\label{A4.eq}
\end{equation}
Then \Eq{A2.eq} becomes the condition that the four real variables
describing a qubit denote a point on the three-sphere $\Sphere{3}$
(a 3-manifold) embedded in $\R^{4}$: 
\begin{equation}
x_{0}^{\ 2}+y_{0}^{\ 2}+x_{1}^{\ 2}+y_{1}^{\ 2}=1\ .\label{A5.eq}
\end{equation}

There is a family of 6 equivalence classes of quadratic maps that
take the remaining 3 degrees of freedom in \Eq{A5.eq} and reduce
them to 2 degrees of freedom by effectively removing $e^{i\phi}$
(``fibering out by the circle $\Sphere{1}$''). The standard form
of this class of maps (``the Hopf fibration'') is 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\\
Z & = & |\alpha_{0}|^{2}-|\alpha_{1}|^{2}=x_{0}^{\ 2}+y_{0}^{\ 2}-x_{1}^{\ 2}-y_{1}^{\ 2}\ .\nonumber 
\end{eqnarray}
These transformed coordinates obey 
\begin{equation}
\|\Vec{X}\|^{2}=X^{2}+Y^{2}+Z^{2}=\left(|\alpha_{0}|^{2}+|\alpha_{1}|^{2}\right)^{2}=1\label{A7.eq}
\end{equation}
and therefore have only two remaining degrees of freedom describing
all possible distinct one-qubit quantum states. In Figure \ref{hopfFibers.fig}
we illustrate schematically the family of circles \textit{each one
of which is collapsed to a point\/} $(\theta,\phi)$ on the surface
$X^{2}+Y^{2}+Z^{2}=1$ by the Hopf map.

%% (We try to list the sources of all computed figures)%% Sources for hopfFibers.fig:  dqc-geom-2012/00mathFigs/Hopf-Fiber.nb
\begin{figure}[htb]
\centerline{ \includegraphics[width=0.48\columnwidth,bb = 0 0 200 100, draft, type=eps]{../../../../Public/Documents/StudyInIUB/Computer Science/201209-12.M800/Manuscripts/hopfS2.pdf}\hspace{0.03in}
\includegraphics[width=0.48\columnwidth,bb = 0 0 200 100, draft, type=eps]{../../../../Public/Documents/StudyInIUB/Computer Science/201209-12.M800/Manuscripts/hopfFibers.pdf}}
\centerline{\hspace{1in} (a) \hfill{}(b) \hspace{1in}} \caption{(a) The sphere represented by \Eq{A7.eq}, which is the irreducible
space of one-qubit states, along with a representative set of points
on the sphere. (b) Representation of the Hopf fibration as a family
of circles (the paths of $e^{i\phi}$), each corresponding to a single
point on the sphere in (a). Points in (a) are color coded corresponding
to circles in (b), e.g., one pole contains the red elliptical circle
that would become an infinite-radius circle in a slightly different
projection, and the opposite pole corresponds to the large perfectly
round red circle at the equator.}
\label{hopfFibers.fig} 
\end{figure}

The resulting manifold is the two-sphere $\Sphere{2}$ (a 2-manifold)
embedded in $\R^{3}$. If we choose one of many possible coordinate
systems describing $\Sphere{3}$ via \Eq{A5.eq} such as 
\begin{eqnarray}
\hspace*{-0.5cm}\left(x_{0},y_{0},x_{1},y_{1}\right) & = & \left(\cos\frac{\theta+\phi}{2}\cos\frac{\psi}{2},\,\sin\frac{\theta+\phi}{2}\cos\frac{\psi}{2},\ \right.\nonumber \\
 &  & \left.\cos\frac{\theta-\phi}{2}\sin\frac{\psi}{2},\ \sin\frac{\theta-\phi}{2}\sin\frac{\psi}{2}\right),\label{A8a.eq}
\end{eqnarray}
where %% $0\leq \theta \leq 2 \pi$,  $-\pi \leq \phi \leq  \pi$, 
$0\leq\psi\leq\pi$, with $0\leq\frac{\theta+\phi}{2}<2\pi$ and $0\leq\frac{\theta-\phi}{2}<2\pi$,
we see that 
\begin{eqnarray}
\left(X,Y,Z\right)=\left(\cos\phi\sin\psi,\,\sin\phi\sin\psi,\,\cos\psi\right)\ .\label{A8b.eq}
\end{eqnarray}
Thus the one-qubit state is independent of $\theta$, and we can choose
$\theta=\phi$ without loss of generality, reducing the form of the
unique one-qubit states to 
\begin{equation}
\ket{\psi_{1}}=e^{i\phi}\cos\frac{\psi}{2}\ket{0}+\sin\frac{\psi}{2}\ket{1}\ ,\label{A9.eq}
\end{equation}
and an irreducible state can be represented as a point on a sphere,
as shown in Figure \ref{blochSphere.fig}(a).

\begin{figure}[thb]
\begin{centering}
\includegraphics[width=0.47\columnwidth,bb = 0 0 200 100, draft, type=eps]{../../../../Public/Documents/StudyInIUB/Computer Science/201209-12.M800/Manuscripts/blochSphere.pdf}
\hspace{0.01in} \includegraphics[width=0.47\columnwidth,bb = 0 0 200 100, draft, type=eps]{../../../../Public/Documents/StudyInIUB/Computer Science/201209-12.M800/Manuscripts/blochSphereArc.pdf} 
\par\end{centering}
\centerline{\hspace{1in} (a) \hfill{}(b) \hspace{1in}} \caption{(a) The conventional Bloch sphere with a unique state represented
by the point at the red sphere. (b) The geodesic shortest-distance
arc connecting two one-qubit quantum states.}
\label{blochSphere.fig} 
\end{figure}

Thus the geometry of a single qubit reduces to transformations among
points on $\Sphere{2}$, which can be parametrized in an infinite
one-parameter family of transformations, one of which is the geodesic
or minimal-length transformation. Explicitly, given two one-qubit
states denoted by points $\Vec{a}$ and $\Vec{b}$ on $\Sphere{2}$,
the shortest rotation carrying the unit normal $\Hat{a}$ to the unit
normal $\Hat{b}$ is the SLERP (spherical linear interpolation) 
\begin{equation}
S(\Hat{a},\Hat{b},t)=\Hat{a}\,\frac{\sin((1-t)\theta)}{\sin\theta}\,+\,\Hat{b}\,\frac{\sin(t\theta)}{\sin\theta}\ ,\label{slerp.eq}
\end{equation}
where $\Hat{a}\cdot\Hat{b}=\cos\theta$. Figure \ref{blochSphere.fig}(b)
illustrates the path traced by a SLERP between two irreducible one-qubit
states on the Bloch sphere. Because states in CQC are defined by infinite
precision real numbers, it is not possible, even in principle, to
make an exact state transition as implied by Figure \ref{blochSphere.fig}(b).
In practice, one has to be content with approximate, typically exponentially
expensive, transitions from state to state.

%% AMR: replaced ', probabilistic, exponentially complex transitions from%% state to state' with the above

\subsection{The $n$-qubit problem}

For $n$ qubits, the irreducible states are encoded in a similar family
of geometric structures known technically as the complex projective
space $\CP{D-1}$. We obtain these structures starting with the $D=2^{n}$
initially unnormalized complex coefficients of the $n$-qubit state
basis 
\begin{eqnarray}
\ket{\Psi} & = & \sum_{i=0}^{D-1}\alpha_{i}\ket{i}\ .\label{N1.eq}
\end{eqnarray}
We then follow the analog of the one-qubit procedure: Conservation
of probability requires that the norm of the vector $\VVec{\alpha}$
be normalized to unity: 
\begin{eqnarray}
\braket{\Psi}{\Psi}=\|\VVec{\alpha}\|^{2} & = & \sum_{i=0}^{D-1}|\alpha_{i}|^{2}\,=\,1\ .\label{probEq1.eq}
\end{eqnarray}
Thus the initial equation for the geometry of a quantum state describes
a \textit{topological sphere\/} $\Sphere{2D-1}$ embedded in $\mathReal^{2D}$.
To see this, remember that we can write the real and imaginary parts
of $\alpha_{i}$ as $\alpha_{i}=x_{i}+iy_{i}$, so 
\begin{eqnarray}
\sum_{i=0}^{D-1}|\alpha_{i}|^{2}=\sum_{i=0}^{D-1}\left(x_{i}^{\ 2}+y_{i}^{\ 2}\right)=1
\end{eqnarray}
describes the locus of a $2D$-dimensional real unit vector in $\mathReal^{2D}$,
which is by definition $\Sphere{2D-1}$, the $(2D-1)$-sphere, with
$D=2^{n}$ for an $n$-qubit state.

This $\Sphere{2D-1}$ in turn is ambiguous up to the usual overall
phase, inducing an $\Sphere{1}$ symmetry action, and identifying
$\Sphere{2D-1}$ as an $\Sphere{1}$ bundle, whose base space is the
$(D-1)$-complex-dimensional projective space $\CP{D-1}$. There are
thus $2D-2$ irreducible real degrees of freedom ($D-1$ complex degrees
of freedom) for a quantum state with a $D$-dimensional basis, $\{\ket{i}\,|\,i=0,\ldots,D-1\}$.

In summary, the full space of a $D=2^{n}$-dimensional $n$-qubit
quantum state, including its overall phase defining its relationship
to other quantum states, is the topological space $\Sphere{2D-1}$.
For an isolated system, the overall phase is not measurable, and eliminating
the phase dependence in turn corresponds to identifying $\Sphere{2D-1}$
as a circle bundle over the base space $\CP{D-1}$, and therefore
$\CP{D-1}=\CP{2^{n}-1}$ defines the $2D-2$ intrinsic, irreducible,
degrees of freedom of the isolated $n$-qubit state's dynamics. In
mathematical notation, this would be written

%% From Wiki: Hopf-fibration: S^1\hookrightarrow S^3 \rightarrow S^2
\[
\Sphere{1}\hookrightarrow\Sphere{2D-1}\rightarrow\CP{D-1}\ ,
\]
with $D=2^{n}$ as usual. For $n=1$, the single qubit, we have $2^{n}-1=2-1=1$,
and the base space of the circle bundle is $\CP{1}=\Sphere{2}$, the
usual Bloch sphere. Note that only for $n=1$ is this actually a sphere-like
geometry due to an accident of low-dimensional topology.

\subsection{Explicit $n$-qubit generalization of the Hopf fibration construction}

For one qubit, we could easily solve the problem of reducing the full
unit-norm space to its irreducible components $\Vec{X}=(X,Y,Z)$ characterizing
the Bloch sphere. We have just argued that essentially the same process
is possible for $n$-qubits: in the abstract argument, we simply identify
the family of coefficients $\{\alpha_{i}\}$ as being the same if
they differ only by an overall phase $e^{i\phi}$. However, in practice
this is not a construction that is easy to realize in a practical
computation. We now outline an explicit algorithm for accomplishing
the reduction to the irreducible $n$-qubit state space $\CP{D-1}$;
this construction will turn out to be useful for the validation of
our discrete results to follow below.

We begin by noting that a natural quantity characterizing an $n$-qubit
system is its \textit{density matrix\/}, $\rho=\left[\alpha_{i}\alpha_{j}^{*}\right]$,
or 
\begin{equation}
\rho=\left[\begin{array}{cccc}
|\alpha_{0}|^{2} & \alpha_{0}\alpha_{1}^{*} & \cdots & \alpha_{0}\alpha_{D-1}^{*}\\
\alpha_{1}\alpha_{0}^{*} & |\alpha_{1}|^{2} & \cdots & \alpha_{1}\alpha_{D-1}^{*}\\
\vdots & \vdots & \ddots & \vdots\\
\alpha_{D-1}\alpha_{0}^{*} & \cdots & \alpha_{D-1}\alpha_{D-2}^{*} & |\alpha_{D-1}|^{2}
\end{array}\right]\ .\label{nqDensity.eq}
\end{equation}

We can now use the complex generalization of the classical Veronese
coordinate system for projective geometry to remove the overall phase
ambiguity $e^{i\phi}$ from the $n$-qubit states. If we take a particular
weighting of the elements of the density matrix $\rho$, we can construct
a \textit{unit vector\/} of real dimension $D^{2}$ with the form:
\begin{equation}
\Vec{X}=(|\alpha_{i}|^{2},\ldots,\sqrt{2}\,\mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\,\mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots)\ ,\label{nqCoords.eq}
\end{equation}
where 
\begin{eqnarray}
\Vec{X}\cdot\Vec{X}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
This construction gives an explicit embedding of the $(D-1)$-dimensional
complex, or $(2D-2)=(2^{n+1}-2)$-dimensional real, object in a real
space of dimension $D^{2}=2^{2n}$. However, this is somewhat subtle
because the vector is of unit length, so technically the embedding
space is a sphere of dimension $D^{2}-1=2^{2n}-1$ embedded in $\R^{D^{2}}$;
the one-qubit irreducible states could be represented in a 4D embedding,
but the magnitude of every coordinate would be one; furthermore, the
object embedded in the resulting $\Sphere{3}$ is indeed $\Sphere{2}$
because we can fix one complex coordinate to be unity, and let one
vary, giving a total of two irreducible dimensions. In fact one must
choose \textit{two\/} coordinate patches, one covering one pole of
$\Sphere{2}$ with coordinates 
\begin{eqnarray}
\begin{array}{ccc}
\alpha_{0} & = & 1+0i\\
\alpha_{1} & = & x_{1}+iy_{1}
\end{array}
\end{eqnarray}
and the other patch covering the other pole of $\Sphere{2}$ with
coordinates 
\begin{eqnarray}
\begin{array}{ccc}
\alpha_{0} & = & x_{0}+iy_{0}\\
\alpha_{1} & = & 1+0i\ .
\end{array}
\end{eqnarray}

We finally see that the irreducible $n$-qubit state space $\CP{D-1}$
is described by $D$ projectively equivalent coordinates, one of which
can always be scaled out to leave $(D-1)$ actual (complex) degrees
of freedom. We must choose, in turn, $D$ different local sets of
complex variables defined by taking the value $\alpha_{k}=1$, with
$k=0,\ldots,D-1$, and allowing the remaining $D-1$ complex (or $2D-2$
real) variables to run free. No single set of coordinates will work,
since the submanifold including $\alpha_{k}=0$ is undefined and another
coordinate system must be chosen to cover that coordinate patch. This
is a standard feature of the topology of non-trivial manifolds such
as $\CP{D-1}$ (see any textbook on geometry \cite{Berger}).

\subsection{The geometry of entanglement}

Entanglement may be regarded as one of the main characteristics distinguishing
quantum from classical mechanics. Entanglement involves quantum correlations
such that the measurement outcomes in one subsystem are related to
the measurement outcomes in another one. Within the standard framework,
given a quantum system composed of $n$ qubit subsystems, a pure state
of the total system $\ket{\Psi}$ is said to be entangled if it cannot
be written as a product of states of each subsystem. That is, a state
$\ket{\Psi}$ is entangled if 
\begin{eqnarray}
\ket{\Psi}\ne\ket{\psi_{1}}\otimes\cdots\otimes\ket{\psi_{j}}\otimes\cdots\otimes\ket{\psi_{n}},
\end{eqnarray}
where $\ket{\psi_{j}}$ refers to an arbitrary state of the $j$-th
qubit, and $\otimes$ represents the tensor product. This is equivalent
to saying that if one calculates the reduced density operator $\rho_{j}$
of the $j$th subsystem by tracing out all the other subsystems, $\rho_{j}={\sf tr}_{\{1,\cdots,j-1,j+1,\cdots,n\}}(\rho)$,
with $j=1,\cdots,n$ and $\rho=\ket{\Psi}\bra{\Psi}$, the normalized
state $\ket{\Psi}$, $\langle\Psi|\Psi\rangle=1$, is entangled if
and only if at least one subsystem state is \textit{mixed\/}; i.e.,
${\sf tr}_{j}(\rho_{j}^{2})<1$. For example, consider 
\begin{eqnarray}
\rho_{j}=\frac{1}{2}(\mathds{1}+\sum\limits _{\mu=x,y,z}\langle\sigma_{\mu}^{j}\rangle\ \sigma_{\mu}^{j}),
\end{eqnarray}
where $\sigma_{\mu}^{j}$, $\mu=x,y,z$, are the Pauli operators acting
on the $j$-th spin, 
\begin{equation}
\sigma_{\mu}^{j}=\overbrace{\mathds{1}\otimes\mathds{1}\otimes\cdots\otimes\underbrace{\sigma_{\mu}}_{j^{th}\ \mbox{factor}}\otimes\cdots\otimes\mathds{1}}^{n\ \mbox{factors}}\;,\label{pauli3}
\end{equation}
with 
\begin{eqnarray}
\hspace*{-0.5cm}\sigma_{x}=\left(\begin{array}{cc}
0 & 1\\
1 & 0
\end{array}\right),\sigma_{y}=\left(\begin{array}{cc}
0 & -i\\
i & 0
\end{array}\right),\sigma_{z}=\left(\begin{array}{cc}
1 & 0\\
0 & -1
\end{array}\right),
\end{eqnarray}
and $\langle\sigma_{\mu}^{j}\rangle=\langle\Psi|\sigma_{\mu}^{j}|\Psi\rangle$
denotes the corresponding expectation value. The vectors ${\bf X}_{j}=(\langle\sigma_{x}^{j}\rangle,\langle\sigma_{y}^{j}\rangle,\langle\sigma_{y}^{j}\rangle)$,
${\bf X}_{j}\in{\mathbb{R}}^{3}$, allow a geometric representation
of each reduced state in ${\mathbb{R}}^{3}$, satisfying $0\le\|{\bf X}_{j}\|\le1$.
Since ${\sf tr}_{j}(\rho_{j}^{2})=\frac{1}{2}(1+\|{\bf X}_{j}\|^{2})$,
the state $\ket{\Psi}$ is entangled if $\|{\bf X}_{j}\|<1$ for at
least one $j$, represented by a point {\em inside} the corresponding
local Bloch sphere. One may therefore consider $\ket{\Psi}$ to be
maximally entangled if $\|{\bf X}_{j}\|=0$ for all $j$. On the other
hand, the state $\ket{\Psi}$ is unentangled (i.e., a product state)
if $\|{\bf X}_{j}\|=1$ for all $j$, corresponding to points lying
on the surface of the Bloch sphere.

A natural geometric measure of multipartite entanglement is obtained
by defining the {\em purity of a state relative to a set of observables}
\cite{Barnum2003,Barnum2004}. If the set is chosen to be the set
of {\em all local observables}, i.e., corresponding to each of
the subsystems that compose the actual system, one recovers the standard
notion of entanglement for multipartite systems. For example, if the
system consists of $n$ qubits, we obtain a measure of conventional
entanglement by calculating the purity relative to the set $\fh=\{\sigma_{x}^{1},\sigma_{y}^{1},\sigma_{z}^{1},\ldots,\sigma_{x}^{n},\sigma_{y}^{n},\sigma_{z}^{n}\}$,
\begin{equation}
P_{\fh}=\frac{1}{n}\sum\limits _{j=1}^{n}\sum\limits _{\mu=x,y,z}\langle\sigma_{\mu}^{j}\rangle^{2}\:,\ \ 0\leq P_{\fh}\leq1\ .\label{purityMeasure.eq}
\end{equation}
Since $\fh$ is a semi-simple Lie algebra, its generalized unentangled
states are the generalized coherent states obtained by applying any
group operation to a reference state such as $\ket{\sf0}=\ket{0}\otimes\cdots\otimes\ket{0}$.
For the algebra $\fh$ of local observables, such group operations
are simply local rotations on each qubit. In other words, the group
orbit describing the generalized coherent states of $\fh$ comprises
all the product states of the form $\ket{\Psi}=\ket{\psi_{1}}\otimes\cdots\otimes\ket{\psi_{n}}$,
which have maximum purity (i.e., $P_{\fh}=1$). Other states such
as the Greenberger-Horne-Zeilinger state $\ket{\Psi}=\ket{{\sf GHZ}_{n}}=\frac{1}{\sqrt{2}}(\ket{0}\otimes\cdots\otimes\ket{0}+\ket{1}\otimes\cdots\otimes\ket{1})$
are (maximally) entangled relative to the set of local observables
(i.e., $P_{\fh}=0$).

Different entanglement measures are obtained when a set $\fh$ different
from the local observables is chosen. An obvious example, in particular,
is given by the set of all observables. In this case, the purity takes
its maximum value independently of the pure quantum state \cite{Barnum2003,Barnum2004},
expressing the fact that any state is a generalized coherent state
of the Lie algebra of all observables.
\end{comment}

@@

\section{Quantum Contextuality}

@@

The characteristic of a field is the least positive integer~$m$
such that $m=1+1+1+\cdots+1=0$, and if no such $m$ exists we say
that the field has characteristic zero (which is the case for $\mathbb{R}$
for example). It turns out that if the characteristic is non-zero,
it must be a prime~$p$. For every prime~$p$ and positive integer~$r$
there is a finite field~$\Fpx{p^{r}}$ of size $q=p^{r}$ and characteristic~$p$,
which is unique up to field isomorphism \cite{Artin1991,DummitFoote2004}.
The exponent $r$ is known as the \emph{degree} of the field over
its prime subfield\footnote{ Fields $\ff{q}$ where $q$ is a power of a prime $p$, i.e., $q=p^{r}$,
are known as Galois fields.}~\cite{GT.ref}. If the characteristic $p$ is an arbitrary prime
number, we call the field \emph{unrestricted}.

@@

Recently, Schumacher and Westmoreland~\cite{Schumacher2012-SCHMQT,SchumacherWestmoreland2010}
and Chang et al.~\cite{doi:10.1142/S0217984913500644,1751-8121-47-40-405304}
defined versions of quantum theory over \textit{unrestricted} finite
fields, which they call modal quantum theories (MQT) or Galois field
quantum theories. Such theories retain several key quantum characteristics
including notions of superposition, interference, entanglement, and
mixed states, along with time evolution using invertible linear operators,
complementarity of incompatible observables, exclusion of local hidden
variable theories, impossibility of cloning quantum states, and the
presence of natural counterparts of quantum information protocols
such as superdense coding and teleportation. These modal theories
are obtained by collapsing the Hilbert space structure over the field
of complex numbers to that of a vector space over an \emph{unrestricted}
finite field. In the resulting structure, all non-zero vectors represent
valid quantum states, and the evolution of a closed quantum system
is described by \emph{arbitrary} invertible linear maps.

Specifically, consider a one-qubit system with basis vectors $\ket{0}$
and~$\ket{1}$. In conventional quantum theory, there exists an infinite
number of states for a qubit of the form $\alpha_{0}\ket{0}+\alpha_{1}\ket{1}$,
with $\alpha_{0}$ and $\alpha_{1}$ elements of the underlying field
of complex numbers subject to the normalization condition $|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=1$.
Moving to a finite field immediately limits the set of possible states
as the coefficients~$\alpha_{0}$ and~$\alpha_{1}$ are now drawn
from a finite set. In particular, in the field $\ff{2}=\{0,1\}$ of
booleans, there are exactly four possible vectors: the zero vector,
the vector~$\ket{0}$, the vector $\ket{1}$, and the vector $\ket{0}+\ket{1}=\ket{+}$,
where $\left\{ \ket{0},\ket{1},\ket{+}\right\} $ is denoted as $\ffzx{2}{2}$.
Since the zero vector is considered non-physical, a one-qubit system
can be in one of only three states. The dynamics of these one-qubit
states is realized by any invertible linear map, i.e., by any linear
map that is guaranteed never to produce the zero vector from a valid
state. There are exactly $6$ such maps: 
\begin{equation}
\begin{aligned}X_{0} & =\begin{pmatrix}1 & 0\\
0 & 1
\end{pmatrix}\,, & S & =\begin{pmatrix}1 & 0\\
1 & 1
\end{pmatrix}\,, & D_{1} & =\begin{pmatrix}0 & 1\\
1 & 1
\end{pmatrix}\,,\\
X_{1} & =\begin{pmatrix}0 & 1\\
1 & 0
\end{pmatrix}\,, & S^{\dagger} & =\begin{pmatrix}1 & 1\\
0 & 1
\end{pmatrix}\,, & D_{2} & =\begin{pmatrix}1 & 1\\
1 & 0
\end{pmatrix}\,.
\end{aligned}
\end{equation}
This set of maps is clearly quite impoverished compared to the full
set of one-qubit unitary maps in conventional quantum theory. In particular,
it does not include the Hadamard transformation. However, this set
also includes non-unitary maps such as $S$ and $S^{\dagger}$ that
are not allowed in conventional quantum computation.
